---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
In order to have a look at the data, we first have to unzip it and load it:
```{r loaddata, echo=TRUE}
unzip("activity.zip", exdir = ".")
activity <- read.csv("activity.csv",header = TRUE)
```

In order to be able to process the data further, we first need to transform the data into a suitable format:
```{r preprocess, echo=TRUE}
# Create an empty vector with length = amount of days present in dataset
daysum <- vector(mode="numeric",length=length(unique(activity$date)))

# For each date, calculate the sum of all steps taken that day, ignoring NAs
for(i in 1:length(unique(activity$date))){
    daysum[i] <- sum(activity$steps[activity$date==unique(activity$date)[i]], na.rm=TRUE)
}
```


## What is the mean total number of steps taken per day?

Here we'll create a histogram of the total number of steps taken each day:

```{r histogram, echo=TRUE}
hist(daysum, main="Histogram of total number of steps taken per day",
     xlab="Number of steps", ylab="Frequency",
     col="darkmagenta")
```

We will now calculate the **mean** and **median** total number of steps taken per day:
```{r meanmedian, echo=TRUE}
meansteps <- mean(daysum)
mediansteps <- median(daysum)
```

The mean total amount of steps taken per day is `r meansteps`, the median is `r mediansteps`

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
